---
title: "STAT 4410/8416 Homework 5"
author: "Gerjol, Nicholas"
date: "Due on Nov 30, 2021"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align='center', dpi=100, message=FALSE, warning=FALSE, cache=TRUE)
output <- opts_knit$get("rmarkdown.pandoc.to")
if(!is.null(output)) {
if (output=="html") opts_chunk$set(out.width = '400px') else
  opts_chunk$set(out.width='.6\\linewidth')
}
```

**1.** **Working with Databases:** Please follow the instruction below before answering the questions:

  • Install the package sqldf using install.packages('sqldf')
  • Import the library using library('sqldf')
  • Read the file https://raw.githubusercontent.com/dsindy/kaggle-titanic/master/data/train.csv and store it in an object called titanic
  
We can now start writing SQL Script using SQLDF library right inside R. See example below:
```{r}
library(sqldf)
library(data.table)
titanic <- fread("https://raw.githubusercontent.com/dsindy/kaggle-titanic/master/data/train.csv")
sqldf("SELECT passengerid, name, sex
FROM titanic
limit 5", drv="SQLite")
```
Answer the following questions. Write SQL Script where applicable.
   a) What does the following command do in MySQL?
         i) show databases;
```{r}
#Show data bases lists all the databases accessible by MySQL
```
         
         ii) show tables;
```{r}
#Shows the data tables inside the databases accessible by MySQL
```
         
Write SQL script to answer the following questions based on titanic data. Display the results of your
script.
i. What is the average age of passengers who survived? Group the data by Sex. Display only the
column Sex, AverageAge
```{r}
sqldf("select sex, avg(age) as AverageAge from titanic where Survived = 1 group by sex
", drv="SQLite")
```


ii. What is the percentage of passengers who survived in each Passenger Class or Pclass? Group
the data by Sex. Display Pclass, Sex, percentage value.
```{r}
#sqldf("select sex, Pclass, sum(Survived) where Survived = 1 / sum(survived) as percentagevalue from titanic group by sex, Pclass", drv="SQLite")
```
iii. What is the average age of all the passenger (survived and not survived)? Group the data by
Pclass, Sex, Survived. After that use ggplot to generate a line plot to show average age vs
pclass, facet by sex and color it by survived.
```{r}
library(ggplot2)
sqldf("select avg(age) as AverageAge from titanic
", drv="SQLite")
plotdat <- sqldf("select Pclass, sex, Survived, avg(age) as AverageAge from titanic group by Pclass, Sex, Survived
", drv="SQLite")
plotdat
g <- ggplot(plotdat, aes(x=Pclass, y = AverageAge), color = Survived) + geom_line() + facet_wrap(~Sex)
g
```
iv. What is the name, age, sex and pclass of the 5 oldest and 5 youngest persons who died?
```{r}
info <- sqldf("select name, age, sex, Pclass from titanic where survived = 0  order by age desc limit 5
", drv="SQLite")
info
info <- sqldf("select name, age, sex, Pclass from titanic where survived = 0  order by age asc limit 5
", drv="SQLite")
info

```

v. On average which Passenger Class is more expensive?
```{r}
info <- sqldf("select Pclass, avg(fare) as avgfare from titanic group by Pclass order by avgfare desc limit 1
", drv="SQLite")
info
```

c. Notice the following R codes and explain what it is doing.

```{r}
library(RSQLite)
conn <- dbConnect(RSQLite::SQLite(), "titanicDB")
dbWriteTable(conn, name = "titanic", value = titanic, overwrite=TRUE)
dbListTables(conn)
#The code is connecting to the titanic data base then renaming the titanicdb as "titanic"
```
d. Use package dplyr to obtain the same result as you did in 1b(iii) above. For this use the connection
string conn and the function tbl(). Store the result in an object called meanAge.
```{r}
library(dplyr)
#conDplyr = src_mysql(dbname = "trainingDB", user = "training", password = "training123", host = "localhost")
# meanAge <- conDplyr %>%
#   tbl("titanic") %>%
#   select(avg(age)) %>%
#   collect() 
#I commented this because I'm still unable to get mysql on my machine so localhost connects to nothing

```

e. Show the SQL query to create meanAge in 1(d) using the fiunction show_query()

2. Extracting twitter data: In this problem we would like to extract data from twitter. For this refer
to the documentation in the e following link.
https://github.com/geoffjentry/twitteR/
a. Twitter API: Set up twitter API using any of the following methods. Make sure you installed all the
packages as mentioned in the class.
Method 2: If you don’t like creating an account with twitter and going through all the trouble, you can use
my keys (ssh, don’t tell anyone). For this download the hw5-twitter-auth file from blackboard and load it
as follows.
```{r}
load("hw5-twitter-auth")
library(twitteR)
setup_twitter_oauth(api_key,api_secret,access_token, access_token_secret)

```
b. Now search twitter messages for “data science job”. Display few job informations.
```{r}
dsTweet <- searchTwitter("data science job")
dfTweet <- twListToDF(dsTweet)
tweetText <- dfTweet$text
head(tweetText)
```

c. Search 300 tweets using the hash tag #chess and save them in an object called rTweets. Show the
top 7 sources of tweets (such as android or iphone) in a ordered bar plot.
```{r}
library(ggplot2)
rTweets <- searchTwitter("#chess", n=300)
sources <- sapply(rTweets, function(x) x$getStatusSource())
sources <- gsub("</a>", "", sources)
sources <- strsplit(sources, ">")
sources <- sapply(sources, function(x) ifelse(length(x) > 1, x[2], x[1]))
source_table = table(sources)
df <- data.frame(names(source_table),source_table)
ggplot(df[df$Freq>8,], aes(reorder(sources,Freq),Freq)) +
  geom_bar(stat="identity") + coord_flip()
```

d. Notice that the object rTweets is a list. Convert it into a data frame using function twListToDF and
store it in an object called dTweets. Display some data from dTweets.
```{r}
dTweets <- twListToDF(rTweets)
head(dTweets)
```

e. dTweets has a column showing the time the tweet was created. Generate a plot showing number of
tweets on each of the hours. Add a smooth line overlaid on your plot.
```{r}
library(lubridate)
library(data.table)
dTweets$created <- ceiling_date(dTweets$created, unit="hour") 
dTweets$num_tweets <- 1
df = subset(dTweets, select = c(created, num_tweets))
test <- aggregate(. ~ created, df, sum)

g <- ggplot(test, aes(x=created, y=num_tweets)) + geom_line()
g
```

f. Arrange the dataframe dTweets based on the retweetCount. While doing this select only columns
text, screenName, retweetCount. Store the data in a object called mostTweets. Display five texts
that are most retweeted.
```{r}
head(dTweets)
dTweets <- as.data.table(dTweets)
testsort <- dTweets[order(-retweetCount),]
head(testsort)
mostTweets <- subset(testsort, select = c(text, screenName, retweetCount))
head(mostTweets, 5)
```

g. Generate a bar chart showing top 15 screen names and count of retweets from mostTweets. Order the
bars based on the retweet counts.
```{r}
topTweets <- head(mostTweets, 15)
bartweets <- subset(topTweets, select = c( screenName, retweetCount))
b <- ggplot(bartweets, aes(x=screenName, y=retweetCount)) + geom_bar(stat = "identity") + coord_flip()
b
```



